---
title: "BISON Tutorial"
output:
  html_document: default
  pdf_document: default
date: '2025-02-12'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

### Introduction
BISON is a Bayesian bi-clustering framework with feature selection for spatially resolved transcriptomic data. It detects informative genes with heterogeneity among spots and clusters spots and informative genes simultaneously. 

### Installation
Load the reqired packages and the source files.

```{r cars, results = 'markup'}
suppressMessages(library(SingleCellExperiment))
suppressMessages(library(scran))
suppressMessages(library(scater))
suppressMessages(library(Matrix))
suppressMessages(library(purrr))
suppressMessages(library(Rcpp))
suppressMessages(library(RcppArmadillo))
source("R/utils.R")
source("R/main.R")
```

### Load MOB data

The current SRT data is stored in a ``SingleCellExperiment`` object. The preprocess 
of SRT data is to keep 1000 highly variable genes (HVGs). 

```{r pressure, results = 'markup'}
load("application/data/MOB_sce_filter.RData")

n_gene = 1000
K = 4
L = 3

sce <- scater::logNormCounts(sce1)
dec <- scran::modelGeneVar(sce, assay.type = "logcounts")
top <- scran::getTopHVGs(dec, n = n_gene)
sce = sce[top, ] # Keep n_gene HVGs in the following analysis
sce
```
Construct adjacent matrix and neighborhood information for spots.
```{r, results = 'markup'}
Adj = find_neighbors(sce, "ST", "lattice")
neighbors = find_neighbor_index(Adj, "ST")
head(neighbors)
```
Construct input count matrix and calculate size factor and gene factor.
```{r}
sizefactor = sce$sizefactor
genefactor = rowData(sce)$genefactor
count_mat = assay(sce, "counts")
P = dim(count_mat)[1]
N = dim(count_mat)[2]

s_mat = matrix(rep(sizefactor, each = P), P, N)
g_mat = matrix(rep(genefactor, N), P, N)
sg = s_mat * g_mat

dim(count_mat)
dim(sg)
```
### Run BNPSpace model
The result is list containing predicted gene labels, predicted spot labels and full MCMC iterations. 
```{r}
source("R/main.R")
timestart = Sys.time()
result = BISON(count_mat = count_mat, sg = sg, neighbors = neighbors, 
               K = K, R = R, f = 1, n_iters = 50, seed = 1)
timeend = Sys.time()
timerun = timeend - timestart
print(timerun)
```

Obtain the etimation label for genes and spots.
```{r}
pred_spot_label = result$pred_spot_label
pred_gene_label = result$pred_gene_label
suppressMessages(library(flexclust))
spot_ARI = randIndex(table(pred_spot_label[!is.na(sce$label)], sce$label[!is.na(sce$label)]))
spot_ARI
```
### Plot
```{r, eval=FALSE}
count_mat_normalize = count_mat
for(i in 1:dim(count_mat)[1]){
  count_mat_normalize[i,] = normalize(count_mat[i, ], mode = "standard")
  
}
loc_sce = cbind(colData(sce1)$col, colData(sce1)$row)
gene_group = paste0("Pattern ", pred_gene_label)
library(ggplot2)
library(ComplexHeatmap)
lgd_list <- list(title = "Relative expression level",
                 title_position = "leftcenter-rot",
                 legend_height = unit(5, "cm"))

mob_layer = pred_spot_label
mob_expression = count_mat_normalize
st_color = c("#1F77B4FF", "#FF7F0EFF", "#D62728FF", "#2CA02CFF")

plot_heatmap = Heatmap(mob_expression, 
        row_split = factor(gene_group, levels = paste0("Pattern ", c(0, 1,2,3))),
        
        cluster_rows = FALSE,
        cluster_row_slices = FALSE,
        show_row_dend = FALSE,
        row_title_rot = 0,
        show_row_names = FALSE,
        
        column_split = factor(mob_layer),
        cluster_columns = TRUE,
        cluster_column_slices = FALSE,
        show_column_dend = FALSE,
        row_names_rot = 0,
        show_column_names = FALSE,
        top_annotation = HeatmapAnnotation(foo = anno_block(gp = gpar(fill = st_color))),
        
        heatmap_legend_param = lgd_list,
        
        use_raster = TRUE,
        raster_quality = 4
)
gb = grid.grabExpr(draw(plot_heatmap))
ggsave(gb, filename = "MOB_Heatmap.pdf", width = 12, height = 9)
### The heatmap are stored in folder "application"
```
The computational environment is below. 
```{r}
sessionInfo()
```

